Why does female meiosis allow the opportunity for exploitation by selfish sperm? BETTER TITLE?

### 2013-09-22 writeup_and_notes/male_dependent_drive.tex
draft of discussion

%Since all alleles within an individual rely on individual survival and reproduction for evolutionary success, most of life in a diploid eukaryotic genome is harmonious.
%However, this harmony is incomplete, as in numerous cases an allele can benefit (in the short or long term) from taking advantage of its host.
Despite the apparent unity of the organism, occasionally alleles can
gain an evolutionary advantage at a cost to  individual fitness
\citep{Burt2006}, often by exploiting meiosis and gametogensis.
%A clear opportunity for this conflict arises during gametogenesis where alternative alleles are in direct competition for representation in the next generation.
Because only one of the four products of female meiosis is transmitted to the egg, female meiosis \yb{is particularly vulnerable to such exploitation}  \citep{Sandler1957,Pardo-ManuelDeVillena2001a}. 
%Female meiosis is particularly unfair, only one of the four products of meiosis becomes an egg, while the other threes are disgarded into polar bodies. 
An allele that biases female meiosis in its favor (i.e. a meiotic driver), may increase in frequency even if this driver entails a pleiotropic fitness cost \citep{Prout1973}, generating a genetic conflict between the success of the driver and organismal fitness.
%Alleles that can segregate into the egg more than half the time in heterozygotes, by exploiting asymmetries, 
%can potentially increase in frequency in the population (i.e. experiencing true meiotic drive).
%If such alleles have fitness consequences they are a source of genetic conflict, 
%and can become balanced in the population if their host in homozygotes 
%out weighs their ability to spread to spread through heterozygotes. 
Meiotic drivers observed in nature (in both plants
\citep{Buckler1999,Fishman2005,Fishman2008}, and animals
\citep{Agulnik1990,Wu2005,Pardo-ManuelDeVillena2001c}) highlight this
conflict -- the selfish benefits of drive and the associated
pleiotropic fitness costs sustain a balanced polymorphism
\citep{Prout1973}, 
and often generate on ongoing evolutionary escalation of drive suppressors and enhancers \citep{Dawe1996,Fishman2008}. 
%Such balanced drive systems can cause the subsequent evolution of linked enhancers of drive
%and of supressors of drive throughout the genome. 
%Indeed, the known polymorphic female meiotic drive systems (OM, maize knobs, IN, mimulus, others?) habor
%a great diversity of enhancers and supressors. 
The threat of meiotic drive to organismal fitness is potentially so
severe that many basic properties of
meiosis and o\"{o}genesis, including the initial genome doubling in
meiosis I \citep{Haig1991}, arrested female meiosis
\citep{Mira1998}, centromere machinery \citep{MalikHenikoff}, and sex differences in the recombination rate \citep{Haig2010,Brandvain2012} 
have perhaps evolved to disrupt meiotic drive and enforce fairness.

@@

%COOP VERSION
%Under a recessive fitness cost model our sperm dependent driver
%actually has far more restrictive conditions to invade the population,
%than a standard meiotic driver (see Figure \ref{Invasion_space}).
%That is because our allele which drives only when it is present in the
%fertilizing sperm is guarenteed to form a homzygote. Thus, 
%to even spread in the population when it is rare, our sperm-dependent
%driver has to overcome the cost it suffers in the homozygote state.

%YB VERSION
When a deleterious recessive green-bearded driver does spread, 
	it spends much time at low frequency, because when rare it is unlikely to meet a complimentary sperm. 
However, once common, a green-bearded driver spreads rapidly to fixation due to its
	positive frequency dependent behavior (Inset in Figure \ref{Invasion_space}).  
%Unlike standard meiotic drive systems, a homozygous cost cannot sustain polymorphism of a green-bearded driver 
%	because it already overcame this cost just to enter the population.
This lack of a balanced polymorphism precludes the evolution of an allele that suppresses this form of meiotic drive.

%COOP VERSION
%When conditions are suitable for such alleles to spread into the
%population, they spread slowly through low frequencies as they rarely
%drive as they are unlikely to be fertilized by a sperm with the right
%allele. Once they become established they spread rapidly to fixation due to their
%positive frequency dependent behavior (Inset in Figure \ref{Invasion_space}). 
%Unlike standard meiotic drive systems 
%these alleles can not become balanced in the population  by homozygote
%cost, as they have already over come this cost just to enter the population.

%YB VERSION
Inclusion of a heterozygous fitness cost (e.g. additive fitness cost) further constrains the evolution of a sperm-dependent driver. 
While a  standard female  driver invades from low frequency when the transmission advantage outweighs the fitness cost (XXX),
	a rare sperm dependent driver pays a fitness cost, but drives only weakly because complementary sperm are uncommon. 
This frequency independent fitness cost, and positive frequency dependent drive generates a bistable system. 
%Therefore, even very weak heterozygous fitness costs
%	prevent the spread of a very rare green-bearded driver, unless its drive is very strong. 
%With an additive cost, the dynamics of a green-bearded driver resembles that of an underdominant mutation (in that the system is bistable). 
Below a cutoff frequency the green-bearded driver deterministically decreases in frequency 
	because it pays the  heterozygous fitness cost  but drives ineffectively (see FigureS\ref{bistable_additive}). 
Above this frequency, the allele increases deterministically because the commonness of complementary sperm 
	make drive effective enough to overcome the associated fitness cost.
This bistability prevents sperm-dependent drivers from invading 	
	reasonably sized populations, and assures that if they do invade they will rapidly fix.

While an the allelic identity of sperm could plausibly influence the outcome of female meiosis, 
	limited gene expression in sperm \citep[e.g.][]{Joseph2004}
	suggest a model where sperm influence female meiosis via expression of the fertilizing male's
	diploid genotype (perhaps due to proteins and RNAs packaged into the sperm).
This adds an additional parameter to the system -- the dominance of the drive enhancer in heterozygous males, 
	when this dominance is high, a narrow sliver of parameter space sustaining a polymorphism emerges (see Figure S\ref{Effect_of_dominance})
 	\yb{presumably because positive frequency dependence diminishes with the dominance of the male effect driver.}
%complicates the dynamics of the system, as the dominance of
 % the genotype of the male effect on female drive must be considered as a parameter. 
%Some circumstances the dominance of the drive effect in
%males can allow the  maintenance of the allele a polymorphism, the parameter range that
%permits polymorphism is very narrow (and quite sensitive to the
%dominance of the male effect, see Figure S\ref{Effect_of_dominance}).

Given the difficulty that green-bearded meiotic drivers have entering the population, the speed at
which they fix if they do,
and the narrow parameter range permitting balanced polymorphisms at such loci, 
it seems very unlikely that such alleles
could drive the evolution of female suppressors of sperm-enabled
female meiotic drive.

@@

XY systems don't care. ZW systems (males ZZ, females ZW). If there is
sex ratio distortion how could this play out?

Finally given a system of a balanced meiotic drive system which avoid
its own ill effects through a sperm-based mechanism
, such as that depicted in Figure XX and XX, female-based modifiers 
can evolve that act to further facilitate the sperm's action in
disrupting driver.

@@

Discussion of general conclusion that females have little reason to evolve supression mechanisms to prevent sperm influence on meiosis. 
General logic that sperm genome has to live in a zygote with consequences of its effect on female meiosis, so
it can not generate too dire a consequence.
I THINK THERE IS A MORE SUBTLE POINT . The sperm has special knowledge
that if it allows drive it will end up in the low fitness homozygote.

>>>>

Despite the apparent unity of the organism, occasionally alleles can
gain an evolutionary advantage at a cost to  individual fitness
\citep{Burt2006}, often by exploiting meiosis and gametogensis.
Because only one of the four products of female meiosis is transmitted to the egg, female meiosis is particularly vulnerable to such exploitation \citep{Sandler1957,Pardo-ManuelDeVillena2001a}. 
An allele that biases female meiosis in its favor (i.e. a meiotic driver), may increase in frequency even if this driver entails a pleiotropic fitness cost \citep{Prout1973}, generating a genetic conflict between the success of the driver and organismal fitness.
Meiotic drivers observed in nature (in both plants
\citep{Buckler1999,Fishman2005,Fishman2008}, and animals
\citep{Agulnik1990,Wu2005,Pardo-ManuelDeVillena2001c}) highlight this
conflict -- the selfish benefits of drive and the associated
pleiotropic fitness costs sustain a balanced polymorphism
\citep{Prout1973}, 
and often generate on ongoing evolutionary escalation of drive suppressors and enhancers \citep{Dawe1996,Fishman2008}. 
The threat of meiotic drive to organismal fitness is potentially so
severe that many basic properties of
meiosis and o\"{o}genesis, including the initial genome doubling in
meiosis I \citep{Haig1991}, arrested female meiosis
\citep{Mira1998}, centromere machinery \citep{MalikHenikoff}, and sex differences in the recombination rate \citep{Haig2010,Brandvain2012} 
have perhaps evolved to disrupt meiotic drive and enforce fairness.

@@

When a deleterious recessive green-bearded driver does spread, 
	it spends much time at low frequency, because when rare it is unlikely to meet a complimentary sperm. 
However, once relatively common, a green-bearded driver spreads rapidly to fixation due to its
	positive frequency dependent behavior (Inset in Figure \ref{Invasion_space}).  
This lack of a balanced polymorphism precludes the evolution of an allele that suppresses this form of meiotic drive.

%YB VERSION
Inclusion of a heterozygous fitness cost (e.g. additive fitness cost) further constrains the evolution of a sperm-dependent driver. 
While a  standard female  driver invades from low frequency when the transmission advantage outweighs the fitness cost (XXX),
	a rare sperm dependent driver pays a fitness cost, but drives only weakly because complementary sperm are uncommon. 
This frequency independent fitness cost, and positive frequency dependent drive generates a bistable system. 
Below a cutoff frequency the green-bearded driver deterministically decreases in frequency 
	because it pays the  heterozygous fitness cost  but drives ineffectively (see FigureS\ref{bistable_additive}). 
Above this frequency, the allele increases deterministically because the commonness of complementary sperm 
	make drive effective enough to overcome the associated fitness cost.
This bistability prevents sperm-dependent drivers from invading 	
	reasonably sized populations, and assures that if they do invade they will rapidly fix.

While an the allelic identity of sperm could plausibly influence the outcome of female meiosis, 
	limited gene expression in sperm \citep[e.g.][]{Joseph2004}
	suggest a model where sperm influence female meiosis via expression of the fertilizing male's
	diploid genotype (perhaps due to proteins and RNAs packaged into the sperm).
This adds an additional parameter to the system -- the dominance of the drive enhancer in heterozygous males, 
	when this dominance is high, a narrow sliver of parameter space sustaining a polymorphism emerges (see Figure S\ref{Effect_of_dominance})
 	\yb{presumably because positive frequency dependence diminishes with the dominance of the male effect driver.}

\gc{Add short paragraph about sex chromosomes here? XY systems don't care. ZW systems (males ZZ, females ZW). If there is
sex ratio distortion how could this play out?}

Given the difficulty that self promiting meiotic drivers have entering the population, the speed at
which they fix if they do,
and the narrow parameter range permitting balanced polymorphisms at such loci, 
it seems very unlikely that such alleles
could drive the evolution of female suppressors of sperm-enabled
female meiotic drive.

@@

Our results provide a plausible answer to why female oogenesis and meiosis has
not evolved to prevent sperm alleles from hijacking . 
Such self-promoting sperm alleles are usually prevented from 
entering the population if they have fitness costs, and are not
maintained as polymorphisms if they do. As such, these type of
exploitation can not drive sustained genetic conflict and so it is unlikely
that female oogenesis and meiosis will evolve to prevent their effect.

Gametogenesis and meiosis are common times for genetic conflict. This
is a consequence of the fact that as the set of alleles that consitute
an individual are in the process of parting ways, 
with only a subset of them being assured transmission.
In constrast, the perspective of the fertilizing sperm's genome is different, it has to live with the consequences of any distortion it causes 
to female meiosis through the entire next life cycle of the organism. 
It is therefore in the interest of the fertilizing sperm's genome to
make a high fitness zygote even if that means alleles in sperm
disrupting other copies of themselves from gaining a 
transmission advantage through female meiosis.

The interests of fertilizing sperm and female are quite well aligned during
syngamy. As noted above there are even opportunities for coevolution
between sperm-based supressors of female meiotic drive and the female genome.
 Thus it seems plausible that aspects of oogenesis may actively 
evolve to facilitate the influence of sperm on the outcome of female meiosis.
In light of that it is interesting to consider whether the phylogenetic variation
in the stage of oogenesis when fertilization takes place reflects not
only species' different reproductive ecologies, but also different
selection pressures shaping the interaction of sperm alleles with
female meioses.

\gc{Other pointsf}\\

@@

I THINK THERE IS A MORE SUBTLE POINT . The sperm has special knowledge
that if it allows drive it will end up in the low fitness homozygote.
